Sequential rank agreement methods for comparison of ranked lists

### 2016-05-26 latex/superranker-manuscript.tex
minor updates

Sequential rank agreement can be generalized to partial lists in the
following way. Let $\Lambda_l\subset X$ be the subset of $d_l$ items
that have been ranked highest in list $l$.  The case where all lists are
censored at the same depth $d$ corresponds to $d_1=\cdots=d_L=d$. For
incomplete/censored lists the rank function becomes
\begin{equation}
  \widetilde R_l(X_p) = \left\{\begin{array}{cl} \{R_l(X_p)\} & \text{ for } X_p\in \Lambda_l \\
      \{d_l+1,\dots,P\} & \text{ for } X_p \not\in \Lambda_l\end{array}\right.
\end{equation}
where we only know that the rank for the unobserved items in list $l$
must be larger than the largest rank observed in that list.

@@

The first reference hypothesis is
\begin{eqnarray*}
H_0  & : &  \text{The list rankings correspond to complete randomly}\\
       &  & \text{permuted lists},
\end{eqnarray*}
which not only assumes that there is no information in the data on
which the rankings are based but also that the methods are completely
independent.

@@

\section{Asymptotics}\label{sec:chgpoint}
In many applications it is of interest to estimate a list depths which
satisfies a change point criterion. In this section we investigate the
theoretical properties of our proposed method for this specific
task. As Hall and Schimek \mcomment{REF}, we work with an infinite set
of lists and study the asymptotic behaviour for $L\to\infty$.
However, we do not allow the length of the lists to change with $L$
because in most applications the lengths of the lists are fixed.

@@

We now define the parameter of interest. For any given threshold value $q$, 
a change-point $d^\ast(q)$ in the list agreement is defined by
\begin{equation}
  d^\ast(q) = \min\left\{\argmin_{d\in\{1,\dots,P\}}\textrm{sra}(d)>q, P\right\}.
\end{equation}
The value of $q$ could for example be the agreement of randomly
permuted lists. The corresponding estimated change-point is similarly defined by
\begin{equation}
\widehat{d^\ast_L}(q) = \min\left\{\argmin_{d\in\{1,\dots,P\}}\widehat{\textrm{sra}}_L(d)>q, P \right\}
\end{equation}
The consistency of $\widehat{d^\ast_L}(q)$ then follows by a simple corollary.

\begin{corollary}
For fixed $q$ it follows from Theorem \ref{thm:consistency} that
$\widehat{d^\ast_L}(q) \overset{P}{\longrightarrow} d^\ast(q)$
for $L \rightarrow \infty$.
\label{cor:consistency}
\end{corollary}

\mcomment{Comment on how to apply this in practice (finite sample/bootstrap/etc.) and that $q$ is similar to LoA BA-style.}

@@

We start by defining the following cadlag function
\begin{align}
  N(p; d) = \sum_{s=1}^p 1\left(Q(R(X_s) \leq d) > 0\right), \quad p = 1,\ldots, P
\end{align}
which runs through the list elements in an arbitrary order and counts
how many list elements have a strictly positive probability to have
rank less than or equal to $d$ under the probability measure $Q$.
Note that the cardinality of the set $S(d)$ in equation
(\ref{eq:sumset}) is equal to the counting process evaluated at the last element:
\begin{equation*}
  |S(d)| = N(P;d).
\end{equation*}

The empirical counterpart of the counting process is given by
\begin{align}
  \widehat{N}_L(p; d) &= \sum_{s=1}^p 1\left( \frac{1}{L} \sum_{l=1}^L 1(R_l(X_s) \leq d) > 0\right)\\
        &= \sum_{s=1}^p 1\left(\widehat{Q}_L(R(X_s) \leq d) > 0\right)\nn.
\end{align}
The joint law of the sequence
$\{\widehat{N}_L(p;d)\}_{1\leq p \leq P}$ is completely determined by
the finite set of the jump times of $N(p;d)$. Since
$\{1(R_l(X_s) \leq d); l=1,\dots, L\}$ consists of independent and
identically Bernoulli distributed variables with expectation
$Q(R(X_s) \leq d)$ it follows from the law of large numbers that
$\widehat{Q}_L(R(X_p) \leq d) \overset{P}{\longrightarrow} Q(R(X_p)
\leq d)$ for every $p$ and $d$ as $L \rightarrow \infty$. 
Therefore, we have
\begin{align}
  \sup_{d \in 1,\ldots,P}\sup_{p \in 1,\ldots, P} \left|\widehat{N}_L(p; d) - N(p; d)\right| = o_P(1).\label{eq:countingConv}
\end{align}

The sequential rank agreement given in equation
(\ref{def:sra}) may therefore be rewritten as the following integral
with respect to $N$
\begin{align}
  \textrm{sra}(d) = \int_{1}^P \frac{A(X_p)}{N(P; d)}\mathrm dN(p; d).
\end{align}
The empirical sequential rank agreement is similarly given by
\begin{align}
  \widehat{\textrm{sra}}_L(d) &= \int_1^P \frac{\widehat{A}_L(X_p)}{\widehat{N}_L(P; d)}\mathrm d \widehat{N}_L(p; d) 
\end{align}
and it follows that
\begin{align}
  \sup_{d \in 1,\ldots,P}\left|\widehat{\textrm{sra}}_L(d) - \textrm{sra}(d)\right| \leq \sup_{d \in 1,\ldots,P}\left|U_L(d)\right| - \sup_{d \in 1,\ldots,P}\left|V_L(d)\right|
\end{align}
where
\begin{align}
  U_L(d) &= \int_1^P\frac{A(X_p)}{N(P;d)} \mathrm d\left(\widehat{N}_L(p; d) - N(p; d)\right)\\
  V_L(d) &=  \int_1^P \left(\frac{\widehat{A}_L(X_p)}{\widehat{N}_L(P; d)} - \frac{A(X_p)}{N(P;d)}\right)\mathrm d\widehat{N}_L(p; d)
\end{align}
The conclusion of the proof therefore follows if each of these two
terms are uniformly of order $o_P(1)$ in $d$.  For the first term we
have that
\begin{align}
  \left|U_L(d)\right| \leq \left(\sup_{p \in 1,\ldots,P}\left|\frac{A(X_p)}{N(P;d)}\right|\right)\left(\widehat{N}_L(p; d) - N(p; d)\right)\biggr\rvert_1^P
\end{align}
where the second factor is $o_P(1)$ uniformly in $d$ by equation
(\ref{eq:countingConv}). The first factor is $O(1)$ since $N(P;d)>0$
uniformly in $d$ and because $A(X_p)\le 2P^2$ uniformly in $p$ \mcomment{double check bound}. It
thus follows that $\left|U_L(d)\right| = o_P(1)$ uniformly in $d$.

Similarly we derive an upper bound for $V_L(d)$ by
\begin{align}
 V_L(d) &\leq \sup_{p \in 1,\ldots,P}\left|\frac{\widehat{A}_L(X_p)}{\widehat{N}_L(P; d)} - \frac{A(X_p)}{N(P;d)}\right| \int_1^P \left|\mathrm d\widehat{N}_L(p; d)\right|\label{eq:secondTerm}
\end{align}
It follows directly from the law of large numbers that
$\widehat{A_L}(X_p) = A(X_p) + o_P(1)$ uniformly in $p$ since
$R_i(X_p)$ in equation (\ref{eq:empDistance}) for $i=1,\ldots,L$ are
independent and $Q$-identically distributed random variables with
finite second moment since $A(X_p)\le 2L^2$). Again by equation
(\ref{eq:countingConv}) it follows that
$\widehat{N}_L(P; d) = N(P;d) + o_P(1)$ uniformly in $d$ and therefore
by the continuous mapping theorem and since $N(P;d)>0$) the first factor
in equation (\ref{eq:secondTerm}) is of order $o_P(1)$ uniformly in
$p$ and $d$. The result then follows by noting that the second factor
is bounded by $P$ uniformly in $d$.
% This completes the proof.

>>>>

Sequential rank agreement can be generalized to partial lists in the
following way. Let $\Lambda_l\subset X$ be the subset of $d_l$ items
that have been ranked highest in list $l$.  The case where all lists are
censored at the same depth $d$ corresponds to $d_1=\cdots=d_L=d$. For
incomplete/censored lists the rank function becomes
\begin{equation}
	\widetilde R_l(X_p) = \begin{cases}
	     \{R_l(X_p)\} & \text{for } X_p\in \Lambda_l\\
	     \{d_l+1,\dots,P\} & \text{for} X_p \not\in \Lambda_l
        \end{cases}
\end{equation}
where we only know that the rank for the unobserved items in list $l$
must be larger than the largest rank observed in that list.

@@

The first reference hypothesis is
\begin{eqnarray}
H_0  & : &  \text{The list rankings correspond to complete randomly}\label{eq:permutationHypothesis}\\
       &  & \text{permuted lists}\nn
\end{eqnarray}
which not only assumes that there is no information in the data on
which the rankings are based but also that the methods are completely
independent.

@@

\section{Asymptotics}\label{sec:chgpoint}
In many applications it is of interest to estimate a list depths which
satisfies a change point criterion. In this section we investigate the
theoretical properties of our proposed method for this specific
task. As in \cite{hall:schi:2012} we work with an infinite set
of lists and study the asymptotic behaviour for $L\to\infty$.
However, we do not allow the length of the lists to change with $L$
because in most applications the lengths of the lists are fixed.

@@

We now define the parameter of interest. For any given deterministic threshold function $q\colon\, 1,\ldots,P \mapsto \mathbb{R}_{\geq 0}$
we define the population superlevel set of the sequential rank agreement as
\begin{align}
  \mathcal{L}(q) = \left\{d : \textrm{sra}(d) \geq q(d)\right\}
\end{align}
The values of $q$ could for example be the limits-of-agreement for randomly
permuted lists as in the null-hypothesis in equation (\ref{eq:permutationHypothesis}).
A change-point $d^\ast(q)$ in the list agreement is then defined by the position
\begin{align}
  d^\ast(q) &= 1\left(|\mathcal{L}(q)| > 0\right)\inf \mathcal{L}(q) + 1\left(|\mathcal{L}(q)| = 0\right)P
\end{align}
corresponding to the first list depth where the sequential rank agreement exceeds the threshold 
if such a position exists. Otherwise, the full list is in agreement according to $q$ and the change-point
is defined as the length of the lists. The empirical superlevel set is similarly defined as
\begin{align}
 \widehat{\mathcal{L}}_L(\widehat{q}_L) = \left\{d : \widehat{\textrm{sra}}_L(d) \geq \widehat{q}_L(d)\right\}
\end{align}
where we allow that the threshold function can depend on the sample size as well. The estimated change-point is therefore
\begin{align}
  \widehat{d^\ast_L}(\widehat{q}_L) &= 1(|\widehat{\mathcal{L}}_L(\widehat{q}_L)| > 0)\inf \widehat{\mathcal{L}}_L(\widehat{q}_L) + 1(|\widehat{\mathcal{L}}_L(\widehat{q}_L)| = 0)P
\end{align}

The consistency of the estimated change-point, $\widehat{d^\ast_L}(\widehat{q}_L)$, follows from Theorem \ref{thm:consistency} by the following corollary. 
\begin{corollary}
Let $\widehat{q}_L$ be a threshold function such that $\left\|\widehat{q}_L - q\right\|_\infty = o_P(1)$ for some limiting
function $q$.  Then, $\widehat{d^\ast_L}(\widehat{q}_L) \overset{P}{\longrightarrow} d^\ast(q)$ for $L \rightarrow \infty$.
\label{cor:consistency}
\end{corollary}

\mcomment{Elaborate more on how to apply this in practice (finite sample/bootstrap/etc.) and that $q$ is similar to LoA BA-style?}

@@

We start by defining the following cadlag function
\begin{align}
  N(p; d) = \sum_{s=1}^p 1\left(Q(R(X_s) \leq d) > 0\right), \quad p = 1,\ldots, P
\end{align}
which runs through the list elements in an arbitrary order and counts
how many list elements have a strictly positive probability to have
rank less than or equal to $d$ under the probability measure $Q$.
Note that the cardinality of the set $S(d)$ in equation
(\ref{eq:sumset}) is equal to the counting process evaluated at the last element: $|S(d)| = N(P;d)$.
The empirical counterpart of the counting process is given by
\begin{align}
  \widehat{N}_L(p; d) &= \sum_{s=1}^p 1\left( \frac{1}{L} \sum_{l=1}^L 1(R_l(X_s) \leq d) > 0\right)\\
        &= \sum_{s=1}^p 1\left(\widehat{Q}_L(R(X_s) \leq d) > 0\right)\nn.
\end{align}
The joint law of the sequence
$\{\widehat{N}_L(p;d)\}_{1\leq p \leq P}$ is completely determined by
the finite set of the jump times of $N(p;d)$. Since
$\{1(R_l(X_s) \leq d); l=1,\dots, L\}$ consists of independent and
identically Bernoulli distributed variables with expectation
$Q(R(X_s) \leq d)$ it follows from the law of large numbers that
$\widehat{Q}_L(R(X_p) \leq d) \overset{P}{\longrightarrow} Q(R(X_p)
\leq d)$ for every $p$ and $d$ as $L \rightarrow \infty$. 
Therefore, we have
\begin{align}
  \sup_{d \in 1,\ldots,P}\sup_{p \in 1,\ldots, P} \left|\widehat{N}_L(p; d) - N(p; d)\right| = o_P(1).\label{eq:countingConv}
\end{align}

The sequential rank agreement given in equation
(\ref{def:sra}) may therefore be rewritten as the following integral
with respect to $N$
\begin{align}
  \textrm{sra}(d) = \int_{1}^P \frac{A(X_p)}{N(P; d)}\mathrm dN(p; d).
\end{align}
The empirical sequential rank agreement is similarly given by
\begin{align}
  \widehat{\textrm{sra}}_L(d) &= \int_1^P \frac{\widehat{A}_L(X_p)}{\widehat{N}_L(P; d)}\mathrm d \widehat{N}_L(p; d) 
\end{align}
and it follows that
\begin{align}
  \sup_{d \in 1,\ldots,P}\left|\widehat{\textrm{sra}}_L(d) - \textrm{sra}(d)\right| \leq \sup_{d \in 1,\ldots,P}\left|U_L(d)\right| - \sup_{d \in 1,\ldots,P}\left|V_L(d)\right|
\end{align}
where
\begin{align}
  U_L(d) &= \int_1^P\frac{A(X_p)}{N(P;d)} \mathrm d\left(\widehat{N}_L(p; d) - N(p; d)\right)\\
  V_L(d) &=  \int_1^P \left(\frac{\widehat{A}_L(X_p)}{\widehat{N}_L(P; d)} - \frac{A(X_p)}{N(P;d)}\right)\mathrm d\widehat{N}_L(p; d)
\end{align}
The conclusion of the proof therefore follows if each of these two
terms are uniformly of order $o_P(1)$ in $d$.  For the first term we
have that
\begin{align}
  \left|U_L(d)\right| \leq \left(\sup_{p \in 1,\ldots,P}\left|\frac{A(X_p)}{N(P;d)}\right|\right)\left(\widehat{N}_L(p; d) - N(p; d)\right)\biggr\rvert_1^P
\end{align}
where the second factor is $o_P(1)$ uniformly in $d$ by equation
(\ref{eq:countingConv}). The first factor is $O(1)$ since $N(P;d)>0$
uniformly in $d$ and because $A(X_p) = O(P^2)$ uniformly in $p$ by the Cauchy-Ekstr{\o}m inequality. It
thus follows that $\left|U_L(d)\right| = o_P(1)$ uniformly in $d$.

Similarly we derive an upper bound for $V_L(d)$ by
\begin{align}
 V_L(d) &\leq \sup_{p \in 1,\ldots,P}\left|\frac{\widehat{A}_L(X_p)}{\widehat{N}_L(P; d)} - \frac{A(X_p)}{N(P;d)}\right| \int_1^P \left|\mathrm d\widehat{N}_L(p; d)\right|\label{eq:secondTerm}
\end{align}
and it follows directly from the law of large numbers that
$\widehat{A_L}(X_p) = A(X_p) + o_P(1)$ uniformly in $p$ since
$R_i(X_p)$ in equation (\ref{eq:empDistance}) for $i=1,\ldots,L$ are
independent and $Q$-identically distributed random variables with
finite second moment. Again, $\widehat{N}_L(P; d) = N(P;d) + o_P(1)$ uniformly in $d$ by equation
(\ref{eq:countingConv}) and $N(P;d)>0$ so by the continuous mapping theorem the first factor
in equation (\ref{eq:secondTerm}) is of order $o_P(1)$ uniformly in
$p$ and $d$. The result then follows by noting that the second factor
is bounded by $P$ uniformly in $d$.
% This completes the proof.
